Neutron Star Mergers and their Impact on Second Generation Star Formation in the Early Universe

### 2023-07-17 NSM_mixing.tex
Making a few edits. Final draft.

% Abstract of the paper
\begin{abstract}
The exact evolution of elements in the universe, from primordial hydrogen and helium to heavier elements like gold and platinum produced via the r-process, is still under scrutiny. The supernova deaths of the very first stars to form in the universe led to the enrichment of their local environments with new metals, and can leave behind neutron stars as remnants. These compact objects can end up in binary systems with other neutron stars, and eventually merge, allowing for the r-process to take place. In this work, we study the scenario where a single neutron star merger (NSM) enriches a halo early in its evolution in order to understand the impact on the second generation of stars and their metal abundances. We perform a suite of high resolution cosmological zoom-in simulations using \enzo{} where we have implemented a new NSM model varying the explosion energy and the delay time. We find that in general, a NSM leads to a significant r-process enhancement in the second generation of stars. A high explosion energy leads to all enhanced r-process stars being highly r-process enhanced, while a lower explosion energy leads to a higher mass fraction of stars being r-process enhanced, but not as many being highly r-process enhanced. When a NSM has a short delay time, there is a higher mass fraction of stars being r-process enhanced, but a smaller mass fraction being highly r-process enhanced. This work represents a stepping stone towards understanding how these events impact their environments and metal abundances of descendant generations of stars.
\end{abstract}

@@

In this work, we are concerned with compact binary mergers, and specifically with two NSs that merge together. Observationally, this has been confirmed. The GW170817 event was detected by LIGO, and along with it came a kilonova, a transient burst of energy powered by radioactive decay, making this event a multimessenger astronomical affair. LIGO detected a NSM with components in the mass range of $1.17 - 1.60 \Ms$ \citep{Abbott17}, and a $\gamma$-ray burst 1.7 seconds after the merger was detected by the Fermi Gamma-ray Burst Monitor \citep{Goldstein17}. This allowed for precise sky localization and a massive follow up effort from telescopes across the electromagnetic spectrum \citep[see][for an extensive review of this follow-up]{Abbott17a}. For the first time, we not only observed a NSM via GWs, but we also confirmed the association between NSMs and $\gamma$-ray bursts. Out of this massive follow-up campaign came analysis of the light curve of the kilonova in the UV, optical, and IR \citep{Drout17, Pian17}. \citet{Drout17} found that this kilonova, was in fact powered by radioactive decay of r-process elements. They estimate that $\sim 0.05 \Ms$ of enriched, r-process material must have been ejected from this event. This was the first tangible evidence that NSMs can produce a significant amount of r-process material.

@@

We use the AMR, hydrodynamical simulation code \enzo{} \citep{Enzo} to run high-resolution proof-of-concept cosmological simulations of a NSM within a zoom-in simulation of a single progenitor halo, and we analyze the simulations with the analysis toolkit \yt{} \citep{yt_full_paper}. We run a total of six simulations, five with a NSM model and one without. The run without the NSM model was run first, and the following five NSM simulations are run based on the non-NSM (original) simulation. The original simulation has a domain width of 1 comoving Mpc and a root grid of $64^{3}$ cells. We use the 12-species (H$_{2}$, H$_{2}^{+}$, H$^{-}$, H, H$^{+}$, He, He$^{+}$, He$^{++}$, e$^{-}$, D, D$^{+}$, and HD) non-equilibrium chemistry model \citep{Abel97, Anninos97}. We consider metal cooling and use the chemistry and cooling library \grackle{} \citep{grackle}. We run this simulation to $z = 10$ and then choose the most massive halo in the box to be the target of a zoom-in simulation. We choose a Lagrangian volume around this halo, and re-initialize the simulation with two more nested grids. The highest refinement level of $l = 14$ has a minimum cell size of $8.6 \times 10^{-2}$ pc and a dark matter mass resolution of $1985 \Ms$. We refine a cell when the baryon or dark matter overdensity exceeds $3 \bar{\rho}_{\mathrm{b, DM}} \times 2^{l (3 + \epsilon)}$ where $l$ is the refinement level and $\epsilon = -0.3$ makes the refinement super-Lagrangian. We also refine around must-refine particles that are initialized in the halo's Lagrangian region, which allows for added refinement in the 8 nearest cells around moving regions up to a level of two. The Jeans length is always resolved by at least four cells. We use the cosmological parameters from the Planck Collaboration (2020) \citep{Planck18_Cosmo}: $\Omega_{\mathrm{M}} = 0.3111$, $\Omega_{\Lambda} = 0.6889$, $\Omega_{\mathrm{DM}} = 0.262125$, $\Omega_{\mathrm{b}} = 0.04897$ and $h = 0.6766$, with the parameters defined as their usual definitions.

@@

Figure \ref{fig:eng_NSM_deltat} shows the r-process metallicity of the Pop II stars that formed within the halo as a function of their creation time relative to the NSM event. We define stars to be ``highly r-process enhanced'' if they have a metallicity of $> 10^{-1} Z_{\mathrm{r}}$ and to be ``r-process enhanced'' if they have a metallicity of $> 10^{-6} Z_{\mathrm{r}}$. At $z = 10$, we find that runs D, Fiducial and E have Pop II mass fractions of 80\%, 80\% and 74\% being r-process enhanced, and Pop II mass fractions of 14\%, 63\%, and 72\% being highly r-process enhanced, respectively. The lowest explosion energy results in a lower percentage of highly r-process enhanced stars, but a similar percentage of r-process enhanced stars compared to the higher explosion energy runs. As the energy increases, all Pop II stars that are enriched by the NSM are highly enhanced. This is due to the large amount of Pop II stars that form at late times near the center of the halo, where the majority of the r-process enhanced gas is contained. Figure \ref{fig:eng_NSM_deltat} also indicates that the Pop II stars that form closer to the NSM tend to have higher r-process metallicity fractions than if they were to form later on. We also see that the higher explosion energies produce Pop II stars with higher r-process metallicities, but less Pop II stars overall.

>>>>

% Abstract of the paper
\begin{abstract}
The exact evolution of elements in the universe, from primordial hydrogen and helium to heavier elements like gold and platinum produced via the r-process, is still under scrutiny. The supernova deaths of the very first stars to form in the universe led to the enrichment of their local environments with new metals, and can leave behind neutron stars as remnants. These compact objects can end up in binary systems with other neutron stars, and eventually merge, allowing for the r-process to take place. In this work, we study the scenario where a single neutron star merger (NSM) enriches a halo early in its evolution in order to understand the impact on the second generation of stars and their metal abundances. We perform a suite of high resolution cosmological zoom-in simulations using \enzo{} where we have implemented a new NSM model varying the explosion energy and the delay time. We find that in general, a NSM leads to a significant r-process enhancement in the second generation of stars. A high explosion energy leads to almost all enhanced r-process stars being highly r-process enhanced, while a lower explosion energy leads to a higher mass fraction of stars being r-process enhanced, but not as many being highly r-process enhanced. When a NSM has a short delay time, there is a higher mass fraction of stars being r-process enhanced, but a smaller mass fraction being highly r-process enhanced compared to longer delay times. This work represents a stepping stone towards understanding how these events impact their environments and metal abundances of descendant generations of stars.
\end{abstract}

@@

In this work, we are concerned with compact binary mergers, and specifically with two NSs that merge together. Observationally, this has been confirmed. The GW170817 event was detected by LIGO, and along with it came a kilonova, a transient burst of energy powered by radioactive decay, making this event a multimessenger astronomical affair. LIGO detected this NSM with components in the mass range of $1.17 - 1.60 \Ms$ \citep{Abbott17}, and a $\gamma$-ray burst 1.7 seconds after the merger was detected by the Fermi Gamma-ray Burst Monitor \citep{Goldstein17}. This allowed for precise sky localization and a massive follow up effort from telescopes across the electromagnetic spectrum \citep[see][for an extensive review of this follow-up]{Abbott17a}. For the first time, we not only observed a NSM via GWs, but we also confirmed the association between NSMs and $\gamma$-ray bursts. Out of this massive follow-up campaign came analysis of the light curve of the kilonova in the UV, optical, and IR \citep{Drout17, Pian17}. \citet{Drout17} found that this kilonova was in fact powered by radioactive decay of r-process elements. They estimate that $\sim 0.05 \Ms$ of enriched, r-process material must have been ejected from this event. This was the first tangible evidence that NSMs can produce a significant amount of r-process material.

@@

We use the AMR, hydrodynamical simulation code \enzo{} \citep{Enzo} to run high-resolution proof-of-concept cosmological simulations of a NSM within a zoom-in simulation of a single progenitor halo, and we analyze the simulations with the analysis toolkit \yt{} \citep{yt_full_paper}. We run a total of six simulations, five with a NSM model and one without. The run without the NSM model was run first, and the following five NSM simulations were run based on the non-NSM (original) simulation. The original simulation has a domain width of 1 comoving Mpc and a root grid of $64^{3}$ cells. We use the 12-species (H$_{2}$, H$_{2}^{+}$, H$^{-}$, H, H$^{+}$, He, He$^{+}$, He$^{++}$, e$^{-}$, D, D$^{+}$, and HD) non-equilibrium chemistry model \citep{Abel97, Anninos97}. We consider metal cooling and use the chemistry and cooling library \grackle{} \citep{grackle}. We run this simulation to $z = 10$ and then choose the most massive halo in the box to be the target of a zoom-in simulation. We choose a Lagrangian volume around this halo, and re-initialize the simulation with two more nested grids. The highest refinement level of $l = 14$ has a minimum cell size of $8.6 \times 10^{-2}$ pc and a dark matter mass resolution of $1985 \Ms$. We refine a cell when the baryon or dark matter overdensity exceeds $3 \bar{\rho}_{\mathrm{b, DM}} \times 2^{l (3 + \epsilon)}$ where $l$ is the refinement level and $\epsilon = -0.3$ makes the refinement super-Lagrangian. We also refine around must-refine particles that are initialized in the halo's Lagrangian region, which allows for added refinement in the 8 nearest cells around moving regions up to a level of two. The Jeans length is always resolved by at least four cells. We use the cosmological parameters from the Planck Collaboration (2020) \citep{Planck18_Cosmo}: $\Omega_{\mathrm{M}} = 0.3111$, $\Omega_{\Lambda} = 0.6889$, $\Omega_{\mathrm{DM}} = 0.262125$, $\Omega_{\mathrm{b}} = 0.04897$ and $h = 0.6766$, with the parameters defined as their usual definitions.

@@

Figure \ref{fig:eng_NSM_deltat} shows the r-process metallicity of the Pop II stars that formed within the halo as a function of their creation time relative to the NSM event. We define stars to be ``highly r-process enhanced'' if they have a metallicity of $> 10^{-1} Z_{\mathrm{r}}$ and to be ``r-process enhanced'' if they have a metallicity of $> 10^{-6} Z_{\mathrm{r}}$. At $z = 10$, we find that runs D, Fiducial and E have Pop II mass fractions of 80\%, 80\% and 74\% being r-process enhanced, and Pop II mass fractions of 14\%, 63\%, and 72\% being highly r-process enhanced, respectively. The lowest explosion energy results in a lower percentage of highly r-process enhanced stars, but a similar percentage of r-process enhanced stars compared to the higher explosion energy runs. As the energy increases, almost all Pop II stars that are enriched by the NSM are highly enhanced. This is due to the large amount of Pop II stars that form at late times near the center of the halo, where the majority of the r-process enhanced gas is contained. Figure \ref{fig:eng_NSM_deltat} also indicates that the Pop II stars that form closer to the NSM tend to have higher r-process metallicity fractions than if they were to form later on. We also see that the higher explosion energies produce Pop II stars with higher r-process metallicities, but less Pop II stars overall.
